Very simplified likelihood

### 2024-02-22 inst/very_simplified_likelihood.Rmd
derivatives and MLE

```{r,label=defmle}

MLE <- function(r,n){
    d <- 1e-6
    optimize(f_vec_arn,c(d,1-d),r=r,n=n,log=TRUE,maximum=TRUE)
}
MLE(0,9)
MLE(3,9)
MLE(3,7)

MLE_analytic <- function(r,n){
uniroot(
  f=function(B){psigamma(B-n+r-1)-psigamma(B+n)+1/(1-B)},
  interval=c(1.0001,1000)
  )
}

```

@@

```{r,label=makejj}
out <- list()
for(n in 1:10){
  out[[n]]  <- rep(NA,n+1)
  for(r in 0:n){
    if(r==0){
      jj <- 1
    } else if(r==n){n
      jj <- 0
    } else {
      jj <- MLE(r,n)$maximum
    }  
    out[[n]][r+1] <- jj
  }
} 
out
```

@@

```{r usehyper3}
n <- 7
a <- seq(from=0,to=1,by=0.01)
fhyper3 <- function(r,n){
  out <- hyper3()
  out['a'] <- 1
  out['b'] <- r
  for(i in (n-r):n){
    out[c(a=1,b=i)] %<>% dec
  }
  return(out)
}

@@

```{r 2samptest}
both <- c(d_ocon,d_gasly)
bothmax <- MLE(both,22)$maximum

d_ocon_max  <- MLE(d_ocon ,22)$maximum
d_gasly_max <- MLE(d_gasly,22)$maximum
f_vec_arn(bothmax,d_ocon ,22,log=TRUE)
f_vec_arn(bothmax,d_gasly,22,log=TRUE)
f_vec_arn(d_ocon_max ,d_ocon ,22,log=TRUE)
f_vec_arn(d_gasly_max,d_gasly,22,log=TRUE)

@@

```{r readoly}
sametest <- function(d1,d2,n){
  both <- c(d1,d2)
  bothmax <- MLE(both,n)$maximum

  d1max <- MLE(d1,n)$maximum
  d2max <- MLE(d2,n)$maximum
  Lambda <- (
  (f_vec_arn(d2max  ,d1,n,log=TRUE)+f_vec_arn(d2max  ,d2,n,log=TRUE)) -
  (f_vec_arn(bothmax,d1,n,log=TRUE)+f_vec_arn(bothmax,d2,n,log=TRUE))
)
  pchisq(-2*Lambda,df=1,lower.tail=FALSE)
}

>>>>

```{r label=defdiff}
fdash <- function(a,r,n,log=TRUE){
	B <- 1/(1-a)
	out <- (1/(B-1) - psigamma(B+n) + psigamma(B+n-r-1))*B^2
	if(!log){out <- out * f_vec_arn(a,r,n,log=FALSE)}
	return(out)
}
```

Now verify `fdash()`:

```{r label=verifydiff}
a <- 0.34
d <- 1e-3
r <- 4
n <- 10
c(
  numerical  = (f_single(a+d/2,r,n,log=TRUE)-f_single(a-d/2,r,n,log=TRUE))/d,
  analytical = fdash(a,r,n)
)

c(
  numerical  = (f_single(a+d/2,r,n,log=FALSE)-f_single(a-d/2,r,n,log=FALSE))/d,
  analytical = fdash(a,r,n,log=FALSE)
)
```

Now use the derivative to find the maximum likelihood point

```{r,label=defmle}
MLE <- function(r,n,give=FALSE){
    d <- 1e-6
     jj <- optimize(f_vec_arn,c(d,1-d),r=r,n=n,log=TRUE,maximum=TRUE)
	 if(!give){jj <- jj$maximum}
	 return(jj)
}

MLEa <- function(r,n,give=FALSE){
	jj <- uniroot(
		f=function(a){fdash(a,r,n)},
		interval=c(0.01,0.99))
	if(!give){jj <- jj$root}
	return(jj)
}

rbind(
c(MLE(1,9),MLEa(1,9)),
c(MLE(3,9),MLEa(3,9)),
c(MLE(3,7),MLEa(3,7))
)
```

@@

```{r,label=makejj}
out <- list()
for(n in 1:10){
  out[[n]]  <- rep(NA,n+1)
  for(r in 0:n){
    if(r==0){
      jj <- 1
    } else if(r==n){n
      jj <- 0
    } else {
      jj <- MLE(r,n)
    }  
    out[[n]][r+1] <- jj
  }
} 
out
```

@@

```{r usehyper3,cache=TRUE}
n <- 7
a <- seq(from=0,to=1,by=0.01)
fhyper3 <- function(r,n){
  out <- hyper3()
  out['a'] <- 1
  out['b'] <- r
  for(i in (n-r):n){
    out[c(a=1,b=i)] %<>% dec
  }
  return(out)
}

@@

```{r 2samptest}
both <- c(d_ocon,d_gasly)
bothmax <- MLE(both,22)

d_ocon_max  <- MLE(d_ocon ,22)
d_gasly_max <- MLE(d_gasly,22)
f_vec_arn(bothmax,d_ocon ,22,log=TRUE)
f_vec_arn(bothmax,d_gasly,22,log=TRUE)
f_vec_arn(d_ocon_max ,d_ocon ,22,log=TRUE)
f_vec_arn(d_gasly_max,d_gasly,22,log=TRUE)

@@

```{r readoly}
sametest <- function(d1,d2,n){
  both <- c(d1,d2)
  bothmax <- MLE(both,n)

  d1max <- MLE(d1,n)
  d2max <- MLE(d2,n)
  Lambda <- (
  (f_vec_arn(d2max  ,d1,n,log=TRUE)+f_vec_arn(d2max  ,d2,n,log=TRUE)) -
  (f_vec_arn(bothmax,d1,n,log=TRUE)+f_vec_arn(bothmax,d2,n,log=TRUE))
)
  pchisq(-2*Lambda,df=1,lower.tail=FALSE)
}
